---
title: "**DOVE3 -- $\\text{\\underline{D}}$urability $\\text{\\underline{O}}$f $\\text{\\underline{V}}$accine $\\text{\\underline{E}}$fficacy 3**"
date: September 21, 2022
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{DOVE3-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
opt <- options()
options(continue="  ", width=70, prompt=" ")
on.exit(options(opt))
library(DOVE3, quietly=TRUE)
```

## Introduction

\textbf{DOVE3} is an R package for estimating the potentially waning long-term effectiveness 
of vaccination and prior infection against COVID-19 outcomes in observational studies (Lin et al., 2022). 
Effects of all exposures (i.e., vaccination and prior infection) are estimated simultaneously under 
a single Cox regression model, allowing the outcome of interest to be a recurrent event.

\textbf{DOVE3} inputs a rectangular data set with the following information:

\begin{itemize}
\item \textbf{Subject ID}: Number that identifies subjects.

\item \textbf{Entry time}: Calendar time when the subject enters the risk set.

\item \textbf{Event Time}: Calendar time when the subject experiences the clinical outcome of interest (e.g., SARS-CoV-2 Infection, hospitalization, or death caused by infection).

\item \textbf{Censoring Time}: Calendar time when the subject moves out of the risk set.

\item \textbf{Vaccination Time}: Calendar time when vaccination takes place.
  
\item \textbf{Vaccination Type}: Categorical variable indicating which vaccine type the subject receives.

\item \textbf{Infection Time}: Calendar time when the subject is infected.

\item \textbf{Infection Type}: Categorical variable indicating which dominant variant the infection is associated with.
  
\item \textbf{Covariates}: Baseline covariates (e.g., priority group, age, gender).
\end{itemize}

\noindent Of note, an arbitrary number of baseline covariates can be included,
and all of the time variables are measured from the start of the study and are 
specified in units of whole days. 

\vspace{.15in}

The primary analysis tool of the package is \textit{dove3()}, which
returns the estimated hazard/rate ratio for each baseline covariate and
the estimated effectiveness in reducing the instantaneous risk.
The standard errors and 95\% confidence intervals are also provided.

\vspace{.15in}

In addition, the package includes three convenience functions:
\textit{outcome()}, which is used to specify the subject id, entry time, event time, and censoring time 
in the model statement of \textit{dove3()}; and \textit{exposure()}, which is used to specify the 
vaccination time, vaccine type, infection time, and infection type in the model statement of \textit{dove3()}.

## Functions

### \textit{outcome()}

This convenience function is used as the left-hand side of a formula object 
for the sole purpose of simplifying the specification of required input variables:
subject id, entry time, event time, and censoring time.
This function is not intended to be used as a stand-alone feature.
For completeness, the function ensures
that the input data obey basic constraints and returns 
the data in a predictable format for use in internal functions.

\vspace{.15in}

The usage is
```{r eval=FALSE}
outcome(subject_id, entry_time, event_time, censor_time)
```
where \texttt{subject.id} is the number that identifies subjects;
\texttt{entry.time} is the calendar time when the subject enters the risk set;
\texttt{event.time} is the calendar time when the subject experiences the clinical outcome of interest
(NA, Inf, or an arbitrary value greater than the censoring time if the subject does not experience an event);
\texttt{censor.time} is the calendar time when the subject moves out of the risk set. 
Note that all times must be provided in units of whole days.

\subsection{\textit{exposure()}}

This convenience function is used as the right-hand side of a formula object 
for the sole purpose of simplifying the specification of required input variables:
vaccination time, vaccine type, infection time, and infection type.
This function is not intended to be used as a stand-alone feature.
For completeness, the function ensures
that the input data obey basic constraints and returns 
the data in a predictable format for use in internal functions.

\vspace{.15in}

The usage is
```{r eval=FALSE}
exposure(vaccine_time, vaccine_type, infection_time, infection_type)
```
where \texttt{vaccine.time} is the calendar time when vaccination takes place 
(NA, Inf, or an arbitrary value greater than the study end time if the subject is not vaccinated during the study period);
\texttt{vaccine.type} is the categorical variable indicating which vaccine type the subject receives 
(an integer between 1 and the total number of vaccine types, with NA or an arbitrary value within this 
range if the subject is not vaccinated during the study period);
\texttt{infection.time} is the calendar time when the subject is infected
(NA, Inf, or an arbitrary value that is larger than the study end time if the subject is not infected during the study period);
\texttt{infection.type} is the categorical variable indicating which dominant variant the infection is associated with 
(an integer between 1 and the total number of variants under investigation, with NA or an arbitrary value within this range if the subject is not infected during the study period).
Note that all times must be provided in units of whole days.

\subsection{\textit{dove3()}}

This function estimates effectiveness of vaccination and prior infection
under the assumption that the log rate/hazard ratio for 
the effect of vaccine and prior infection is a piecewise linear function of time. The value object 
returned contains the estimated hazard/rate ratio for each baseline covariate and the 
estimated effectiveness of vaccine and prior infection in reducing the 
hazard rate, $EE_h(t)$, where $t$ is the
time elapsed since vaccination or infection.
The 95\% confidence intervals for the estimated effectiveness are provided.

\vspace{.15in}

The function call takes the following form:

```{r eval=FALSE}
dove3(formula,
      data,
      vaccine_infection_interaction = FALSE,
      vaccine_knots = NULL,
      vaccine_uninfected_knots = NULL,
      vaccine_infected_knots = NULL,
      prior_infection_knots = NULL,
      related_vaccine_types = NULL,
      last_piece_constant = FALSE,
      reinfection_cutoff = 14,
      plots = TRUE
)
```
where
\begin{itemize}
\item \texttt{formula} is a model statement. See below for further details. 
\item \texttt{data} is a data.frame object containing all required data
as previously described.
\item \texttt{vaccine\_infection\_interaction} is a logical object specifying the interaction between vaccination and prior infection. If TRUE, vaccine effects are allowed to be different between previously uninfected subjects and previously infected subjects; otherwise, average vaccine effects among all subjects are estimated, regardless of the prior infection status.
\item \texttt{vaccine\_knots} is a list object of which the $k$th element specifies the knots of the piecewise linear function for the log rate/hazard ratio of the
$k$th vaccine type.
\item \texttt{vaccine\_uninfected\_knots} is a list object of which the $k$th element specifies the knots of the piecewise linear function for the log rate/hazard ratio of the $k$th vaccine type given no prior infection before vaccination.
\item \texttt{vaccine\_infected\_knots} is a list object of which the $k$th element specifies the knots of the piecewise linear function for the log rate/hazard ratio of the $k$th vaccine type given at least one prior infection before vaccination.
\item \texttt{prior\_infection\_knots} is a list object of which the $k$th element specifies the knots of the piecewise linear function for the log rate/hazard ratio of the $k$th infection type. The first knot should be placed at reinfection\_cutoff.
\item \texttt{related\_vaccine\_types} is a list object of which the element takes the form c(i,j) and imposes a constraint that the slope of the first piece of the piecewise linear function is the same between the $i$th and $j$th vaccine types.
\item \texttt{last\_piece\_constant} is a logical object specifying the effectiveness trend after the last change point. If TRUE, effectiveness is assumed to be constant after the last change point; otherwise, effectiveness is allowed to vary after the last change point.
\item \texttt{reinfection\_cutoff} is a positive scalar object specifying the gap time (in days) that defines distinct infections. If the gap time of two infections is larger than reinfection\_cutoff, they considered as two different infections.
\item \texttt{plots} is a logical object indicating whether graphical forms of the estimated effectiveness curves are to be generated.
\end{itemize}

If vaccine\_knots, vaccine\_uninfected\_knots, or vaccine\_infected\_knots = NULL, knots will be placed at every other month by default. The input of vaccine\_knots will be ignored when vaccine\_infection\_interaction = TRUE. The input of vaccine\_uninfected\_knots and vaccine\_infected\_knots will be ignored if vaccine\_infection\_interaction = FALSE. If prior\_infection\_knots = NULL, a default set of knots will be used, with the first knot placed at reinfection\_cutoff and the subsequent knots placed at every three months.

\vspace{.15in}

The model statement is a formula object. The left side is an
object returned by the \textit{outcome()} function. 
The right-hand side is a combination of
baseline covariates and the previously described \textit{exposure()} function.
Categorical baseline covariates can be specified, and all other categories are compared to the first category.

The \texttt{formula} input takes the following general structure

```{r intCens-usage, eval=FALSE}
outcome(subject_id, entry_time, event_time, censor_time) ~ covariates +
    exposure(vaccine_time, vaccine_type, infection_time, infection_type)
```

where 'subject\_id', 'entry\_time', 'event\_time', 'censor_time', 'covariates', 'vaccine\_time', 
'vaccine\_type', 'infection\_time', and 'infection\_type'
are place holders indicating the data that are to be provided; they should be replaced by the 
appropriate variable names in the header of the input data.

## Examples

To illustrate the call structure and results of \textit{dove3()}, we use the 
dataset provided with the package, exampleData. 
This dataset was simulated to mimics the surveillance data of
vaccination and infection and contains the following observations 
for each of the 8,000 records:

\begin{itemize}
\item \textbf{subject.id}: The subject ID of each record.
\item \textbf{event.time}: The event (infection, hospitalization or death) time in days. Infection is treated as recurrent event.
\item \textbf{censor.time}: The censoring time for each subject in days.
\item \textbf{entry.time}: The entry time in days.
\item \textbf{Vtime}: The time of vaccination/booster in days.
\item \textbf{Vtype}: The type of vaccination/booster.
\item \textbf{infection.time}: The time of infection in days. If the outcome is infection, infection.time is the same as event.time.
\item \textbf{infection.type}: The type of infection.
\item \textbf{age}: A categorical variable of age (<18, 18-34, 35-49, 50-64, >=65).
\item \textbf{gender}: A binary indicator of gender (1 for male and 0 for female).
\item \textbf{priority}: A composite baseline risk score taking values 1-3.
\end{itemize}

\vspace{.15in}

The data can be loaded in the usual way
```{r data-load}
data(exampleData)
```

```{r data-head}
head(exampleData)
```

\vspace{.15in}

Consider the summary statistics
```{r data-summary}
summary(exampleData)
```
We can see that participants were enrolled in the study over a 5-month period 
(119 $\le$ entry.time $\le 271$ days), the follow-up time ended on day 272
(censor.time $\le$ 272 days). In this analysis, we will include in our model statement all three baseline 
covariates, age, gender and priority.

\vspace{.25in}

In the first example, we estimated the effectiveness of the 3 types of vaccination and the 2 types of the prior infection without interaction between them.
The knots are placed as follows:
```{r eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE}
vaccine.knots = list("vac.type1" = c(30),
                     "vac.type2" = c(30,60),
                     "vac.type3" = c(30,60))
prior.infection.knots = list("inf.type1" = c(14),
                             "inf.type2" = c(14))
```
The function call takes the following form
```{r}
formula = outcome(subject.id, entry.time, event.time, censor.time) ~
  age + gender + priority +
  exposure(Vtime, Vtype, infection.time, infection.type)
data = exampleData

result1 = dove3(formula = formula,
                data = data,
                vaccine_knots = vaccine.knots,
                prior_infection_knots = prior.infection.knots)
```

\vspace{.15in}

The function returns a list object with the following information.

\vspace{.1in}

\noindent \textbf{Covariate Effects}: The estimated (log) hazard/rate ratio of each covariate, 
together with the estimated standard error, the $95\%$ confidence 
interval, and the two-sided p-value for testing no covariate effect. 

```{r}
result1$covariates
```

\vspace{.15in}

\noindent \textbf{Estimated Effectiveness}: Element \textbf{\$effectiveness} contains the 
estimated effectiveness for different types of vaccination or prior infection in reducing 
the attack rate at each observed event time, together 
with its standard error and the $95\%$ confidence interval. The $i$th element of result1$effectiveness
corresponds to the $i$th type of vaccination or prior infection.
      
```{r}
head(result1$effectiveness[[1]])
```

\vspace{.15in}

\noindent \textbf{Plots}: Element \textbf{\$plots} contains the graphical depictions of the estimated effectiveness 
that generated by default by \textit{dove3()}. The $i$th element of result1$plots
corresponds to the $i$th type of vaccination or prior infection, which can be regenerated as follows:
      
```{r}
result1$plots[[1]]
```

\vspace{.15in}

In the second example, we estimated the effectiveness of the 3 types of vaccination and the 2 types of the prior infection with interaction between them.
In addition, we assumed the first piece of the piecewise linear function is the same between the first and second vaccine types. 
The knots are placed as follows:

```{r}
vaccine.uninfected.knots = list("vac.noinf.type1" = c(30),
                                "vac.noinf.type2" = c(30,60),
                                "vac.noinf.type3" = c(30,60))

vaccine.infected.knots = list("vac.noinf.type1" = c(30),
                              "vac.noinf.type2" = c(60),
                              "vac.noinf.type3" = c(60))
```
The function call takes the following form
```{r}
result2 = dove3(formula = formula,
                data = data,
                vaccine_infection_interaction = TRUE,
                vaccine_uninfected_knots = vaccine.uninfected.knots,
                vaccine_infected_knots = vaccine.infected.knots,
                prior_infection_knots = prior.infection.knots,
                related_vaccine_types = list(c(1,2)))
```

\vspace{.15in}

The function returns a list object containing the following items. 

\vspace{.1in}

\noindent \textbf{Covariate Effects}: The estimated (log) hazard/rate ratio of each covariate, 
together with the estimated standard error, the $95\%$ confidence 
interval, and the two-sided p-value for testing no covariate effect.
      
```{r idove-constantVE-return-covariates}
result2$covariates
```

\vspace{.15in}
 

\noindent \textbf{Estimated Effectiveness}: Element \textbf{\$effectiveness} contains the 
estimated constant VE, together 
with its standard error and the $95\%$ confidence interval. 
      
```{r}
head(result2$effectiveness[[1]])
```

\vspace{.15in}

\noindent \textbf{Plots}: Element \textbf{\$plots} contains the graphical depictions of the estimated effectiveness:
      
```{r}
result2$plots[[1]]
```

\noindent \textbf{References}

Lin, D-Y, Gu, Y., Xu, Y., et al. Association of Primary and Booster Vaccination and Prior Infection With SARS-CoV-2 Infection and Severe COVID-19 Outcomes. JAMA. Published online September 26, 2022. doi:10.1001/jama.2022.17876

